---
title: "Statistical Inference - Course Project - Part 2 - Basic Inferential Data Analysis"
author: "Ioannis Moros"
date: "01/07/2017"
output:
  html_document: default
  pdf_document: default
---

## R Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r pack}
require(ggplot2)
```

## Introduction

Now in the second portion of the project, we're going to analyze the ToothGrowth 
data in the R datasets package.

## 1. Load the data and perform basic exploratory data analysis. 

```{r load}
my_data <- datasets::ToothGrowth
```

```{r summary}
str(my_data)

summary(my_data)
```

The boxplot of odontoblast length by supplement type

```{r boxplot}
g <- ggplot(my_data, aes(x = supp, y = len))
g + geom_boxplot(aes(group = supp), fill = "salmon") + 
        xlab("supplement type") + 
        ylab("odontoblast length") + 
        ggtitle("Odontoblast length by supplement type")
```

## 2. Hypothesis tests and conclusions.

We want to investigate if there is a significant difference in odontoblast 
length between the 2 delivery methods, orange juice and ascorbic acid. We will 
perform a two-sided test using the paired t-test:

H_0: there is no difference in growth between the delivery methods

H_1: there is a difference

```{r ttest1}
t.test(len ~ supp, data = ToothGrowth, paired = FALSE)
```

The p-value of the t-test is 0.06 > 0.05, so we retain the null hypothesis at the 95%
level, i.e. we conclude that odontoblast growth is not significantly different 
between the two delivery methods, with 95% confidence. We observe that the 
confidence interval (-0.17, 7.57) contains 0, which is enough evidence to retain 
the null.

Now we investigate the effect of different vitamin C dosages.

```{r plot2}
p <- ggplot(my_data, aes(x = as.factor(dose), y = len))
p + geom_boxplot(aes(group = dose), fill = "steel blue") + 
        xlab("dose (mg. per day)") + 
        ylab("odontoblast length") + 
        ggtitle("Odontoblast length by vitamic C dose")
```

It looks like higher dosages of vitamic C are associated with longer odontoblasts. 
Let's test that hypothesis for dose levels 0.5 and 1 mg/day. Again, we use the 
paired t-test, this time for a one-sided test

H_0: odontoblast is the same for vitamin C dose of 0.5 mg/day and 1 mg/day

H_1: odontoblasts are longer for dose of 1 mg/day

```{r ttest2}
t.test(len ~ dose, data = subset(my_data, dose %in% c(1, 0.5)), alt = "less", paired = FALSE)
```

The p-value is practically 0 , so we reject the null for any reasonable
significance level and we conclude that a dosage of 1mg/day is associated with 
longer odontoblasts than a dosage of 0.5 mg/day.

## 3. Assumptions:

1. Independence of observations - each guinea pig was given exactly one dosage
via exactly one delivery method.
2. Normality - odontoblast length is approximately normally distributed for each
group of the independent variables (supp, dose)
3. Homogeneity of variances.